Reduced growth due to belowground sink limitation is not explained by reduced photosynthesis

### 2016-03-24 manuscript/manuscript_pve.Rmd
manuscript edits BM, errors to make rmd run fixed

csl: net_phytologist.csl
bibliography: references2.bib
---

@@

##Seedling growth model
We developed a simple seedling growth model that utilized leaf *A*~n~ rates to allocate daily C assimilate towards biomass production of stems, leaves, fine roots and coarse roots. The model begins with mean initial tissue component biomass (leaf~i~, stem~i~ and root~i~) and a starting leaf area (LA~i~) measured prior to planting. The initial biomass of roots was divided evenly between fine and coarse roots. The daily net biomass production of seedlings (*P*~i~) is then given by

\
(1)
$$P_i = \frac{L~C_\mathrm{day'i}~\sigma_\mathrm{s} - R}{\epsilon_\mathrm{c}}$$

where *L* is total plant leaf area (m^2^), *C*~day,i~ is the predicted daily carbon assimilation (g C m^-2^ d^-1^, explained below), $\sigma$~s~ is a self-shading parameter (explained below), $\epsilon$~c~ is a biomass conversion efficiency parameter (g C g mass^-1^) and *R* is the mass based total respiration of all tissue components (g C d^-1^). Total respiration was calculated as

@@

\
It was further necessary to calculate a self-shading parameter ($\sigma$~s~) when scaling leaf *A*~n~ with total plant leaf area. This was accomplished by utilizing 61 previously digitized *Eucalyptus* seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package [@YplantQMC] in R to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Using the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total *A*~n~, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf. The ratio of total *A*~n~ with self-shading to horizontal leaf was then used to calculate $\sigma$~s~ for each of the 61 digitized seedlings, independently for each of the seven soil volume treatments. Next, the linear relationship between $\sigma$~s~ and total leaf area was determined across digitized seedlings, within each treatment. For the growth model, $\sigma$~s~ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table `r pvetabS`.

@@

\
Total seedling biomass at harvest was significantly different across container volumes (*P* < 0.001) and between container treatments and free seedlings (*P* < 0.001, Table `r pvetab1`). On average, harvested biomass of free seedlings was 84% higher than that of seedlings in containers. Plant biomass was positively correlated with total leaf area across all treatments (*R*^2^ = 0.97, *P* < 0.001). Differences in biomass partitioning to leaves, stems and roots were not different across soil volumes after variation in seedling biomass across treatments was factored in the analysis (Figure `r allocfig`a,b). Across all treatments, the final harvest root:shoot biomass ratio and  was conserved in these seedlings which exhibited a slightly higher shoot than root mass ($\overline{x}$ = 0.904, 95% CI = [0.846,1.119]) and the ratio of leaf to fine root mass was also not different (Figure `r allocfig`c).

@@

\
Decreases in *A*~sat~ occurred at the same time as reductions in height and diameter of seedlings in containers. This observation initially suggests a strong link between growth and an apparent down regulation of *A*~n~. However, there are several possible mechanisms that can explain reduced *A*~n~ in small containers including reduced nutrient content or water availability, or reduced sink strength [@poorter2012pot]. It was therefore necessary to examine each of these factors to determine if the induced belowground sink limitation actually triggered photosynthetic down regulation.

@@

\
As both rooting space and resources were finite in containers, the inability of seedlings to maintain the capacity of the belowground C sink resulted in the buildup of C assimilate in leaves. The feedback inhibition of *A*~n~ from starch accumulation has been proposed, yet it is still not known whether there is a starch threshold that triggers the down-regulation process [@nebauer2011photosynthesis]. Here, declines in *A*~max~ were correlated with higher starch content throughout the experiment. This agrees with a study on a deciduous conifer by Equiza et al. [-@equiza2006photosynthetic] where photosynthetic downregulation from reduced sink strength was correlated with starch content. As starch content in leaves of plants grown in the smallest containers was nearly double that of free seedlings in our study, this suggests the response of *A*~n~ to sink inhibition was regulated by this accumulation, as hypothesized.

##Biomass partitioning under sink limitation
As biomass partitioning is likely controlled by the source and sink strength of all organs [@poorter2012biomass], it was important to determine which tissue components were most affected by the container size treatments. It was necessary to distinguish if growth was affected beyond ontogenetic constraints, by correcting for size, as biomass distribution is strongly size-dependent [@gould1966allometry; @lleonart2000removing]. In this study, there was no significant difference in root, leaf, or stem biomass partitioning with reduced soil volume compared to free seedlings, outside of ontogenetic drift (Figure `r allocfig`a,b). This is a surprising result as shifts in partitioning have been noted specifically for nutrient limitation [@mcconnaughay1999biomass, and references therein]. Surprisingly, a constant ratio of fine root mass to leaf mass was observed across all treatments suggesting a functional partitioning response to optimize resource gain did not occur.

\
The lack of detected shifts in partitioning to fine roots provides evidence against an optimal foraging strategy for seedlings in containers. This could be because lateral root development is affected by inanimate obstacles and avoiding growth towards container walls could improve the efficiency of resource allocation [@falik2005root]. The sensitivity of roots to their own exudates near obstructions may prevent further growth [@semchenko2008foraging]. Here, we found that FRLD was highest in smallest containers suggesting that root restriction likely occurred as simple function of available rooting space. Additionally, physical restriction of root proliferation could have impacted root development and morphology prior to shifts in mass partitioning. Here, increases in SRL were detected in several of the soil volume treatments. This is not surprising as plants in containers have been shown to have different root morphology to field grown plants [@nesmith1998effect]. The poor soil quality used in our experiment and root restriction, however, likely decreased the capacity of this morphological response to increase N uptake.

>>>>

csl: new-phytologist.csl
bibliography: references2.bib
---

@@

##Seedling growth model
We developed a simple seedling growth model that utilized leaf *A*~n~ rates to allocate daily C assimilate towards biomass production of stems, leaves, fine roots and coarse roots. The model begins with mean initial tissue component biomass (leaf~0,i~, stem~0,i~ and root~0,i~) and a starting leaf area (L~0,i~) measured prior to planting. The initial biomass of roots was divided evenly between fine and coarse roots. The daily net biomass production of seedlings (*P*~i~) is then given by

\
(1)
$$P_i = \frac{L~C_\mathrm{day'i}~\sigma - R}{\epsilon}$$

where *L* is total plant leaf area (m^2^), *C*~day,i~ is the predicted daily carbon assimilation (g C m^-2^ d^-1^, explained below), $\sigma$ is a self-shading parameter (explained below), $\epsilon$ is a biomass conversion efficiency parameter (g C g mass^-1^) and *R* is the mass based total respiration of all tissue components (g C d^-1^). Total respiration was calculated as

@@

\
It was further necessary to calculate a self-shading parameter ($\sigma$) when scaling leaf *A*~n~ with total plant leaf area. This was accomplished by utilizing 61 previously digitized *Eucalyptus* seedlings, covering 5 total species which include *E. tereticornis*, from Duursma et al. [-@duursma2012light] to run in 'YplantQMC' package [@YplantQMC] in R to build a 3D plant structure based on digitized metrics of plant allometry and crown structure. Using the same treatment specific physiological parameters listed above, 'YplantQMC' outputs total *A*~n~, using total leaf area, for seedlings assuming self-shading as well as for a full sun large horizontal leaf. The ratio of total *A*~n~ with self-shading to horizontal leaf was then used to calculate $\sigma$ for each of the 61 digitized seedlings, independently for each of the seven soil volume treatments. Next, the linear relationship between $\sigma$ and total leaf area was determined across digitized seedlings, within each treatment. For the growth model, $\sigma$ was then predicted for each daily time step using the previous days cumulative leaf area and this value  was then applied to C~day,i~. All default parameters used in model simulations are reported in Table `r pvetabS`.

@@

\
Total seedling biomass at harvest was significantly different across container volumes (*P* < 0.001) and between container treatments and free seedlings (*P* < 0.001, Table `r pvetab1`). On average, harvested biomass of free seedlings was 84% higher than that of seedlings in containers. Plant biomass was positively correlated with total leaf area across all treatments (*R*^2^ = 0.97, *P* < 0.001). Differences in biomass partitioning to leaves, stems and roots were not different across soil volumes after variation in seedling biomass across treatments was factored in the analysis (Figure `r allocfig`a,b). The final harvest root:shoot biomass ratio and  was conserved across all treatments, with a slightly higher shoot than root mass ($\overline{x}$ = 0.904, 95% CI = [0.846,1.119]). The ratio of leaf to fine root mass was also not different between treatments (Figure `r allocfig`c).

@@

\
Decreases in *A*~sat~ occurred simultaneously with reductions growth of seedlings in containers. This observation initially suggests 
that sink limitation may be driving the down regulation of photosynthesis, and thus growth. However, there are several possible mechanisms that can explain reduced *A*~n~ in small containers including reduced nutrient content or water availability, or reduced sink strength [@poorter2012pot]. It was therefore necessary to examine each of these factors to determine if the induced belowground sink limitation actually triggered photosynthetic down regulation.

@@

\
As both rooting space and resources were finite in containers, the inability of seedlings to maintain the capacity of the belowground C sink resulted in the buildup of C assimilate in leaves. The feedback inhibition of *A*~n~ from starch accumulation has been proposed, yet it is still not known whether there is a starch threshold that triggers the down-regulation process [@nebauer2011photosynthesis]. Here, declines in *A*~max~ were correlated with higher starch content throughout the experiment. These results agree with a study on a deciduous conifer by Equiza et al. [-@equiza2006photosynthetic] where photosynthetic downregulation from reduced sink strength was correlated with starch content. As starch content in leaves of plants grown in the smallest containers was nearly double that of free seedlings in our study, this suggests the response of *A*~n~ to sink inhibition was regulated by this accumulation, as hypothesized.

##Biomass partitioning under sink limitation
As biomass partitioning is likely controlled by the source and sink strength of all organs [@poorter2012biomass], it was important to determine which tissue components were most affected by the container size treatments. It was necessary to distinguish if growth was affected beyond ontogenetic constraints, by correcting for size, as biomass distribution is strongly size-dependent [@gould1966allometry; @lleonart2000removing]. In this study, there was no significant difference in root, leaf, or stem biomass partitioning with reduced soil volume compared to free seedlings, outside of ontogenetic drift (Figure `r allocfig`a,b). This result is unexpected as shifts in partitioning have been noted specifically for nutrient limitation [@mcconnaughay1999biomass, and references therein]. Surprisingly, a constant ratio of fine root mass to leaf mass was observed across all treatments suggesting a functional partitioning response to optimize resource gain did not occur.

\
The lack of detected shifts in partitioning to fine roots provides evidence against an optimal foraging strategy for seedlings in containers. It is instead possible that lateral root development is affected by inanimate obstacles and avoiding growth towards container walls may improve resource gain [@falik2005root]. The sensitivity of roots to their own exudates near obstructions may also prevent further growth [@semchenko2008foraging]. Here, we found that FRLD was highest in smallest containers suggesting that root restriction likely occurred as simple function of available rooting space. Additionally, physical restriction of root proliferation could have impacted root development and morphology prior to shifts in mass partitioning. Here, increases in SRL were detected in several of the soil volume treatments. This is not surprising as plants in containers have been shown to have different root morphology to field grown plants [@nesmith1998effect]. The poor soil quality used in our experiment and root restriction, however, likely decreased the capacity of this morphological response to increase N uptake.
